---
output: github_document
bibliography: vignettes/References.bib
---

<!-- *********** -->
<!-- Note: README.md is generated from README.Rmd.   -->
<!-- Be sure to edit README.Rmd and generate the README.md file by Cmd/Ctl-shift-K -->
<!-- *********** -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# matsbyname

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/matsbyname)](https://cran.r-project.org/package=matsbyname)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/MatthewHeun/matsbyname/workflows/R-CMD-check/badge.svg)](https://github.com/MatthewHeun/matsbyname/actions)
[![Codecov test coverage](https://codecov.io/gh/MatthewHeun/matsbyname/branch/master/graph/badge.svg)](https://codecov.io/gh/MatthewHeun/matsbyname?branch=master)
<!-- badges: end -->

The `R` package `matsbyname` enables 
matrix mathematics wherein operations are performed "by name" and
row and column types are allowed.


## Installation

You can install `matsbyname` from CRAN with:

```{r CRAN-installation, eval = FALSE}
install.packages("matsbyname")
```

You can install a recent development version of `matsbyname` from github with:

```{r gh-installation, eval = FALSE}
# install devtools if not already installed
# install.packages("devtools")
devtools::install_github("MatthewHeun/matsbyname")
# To build vignettes locally, use
devtools::install_github("MatthewHeun/matsbyname", build_vignettes = TRUE)
```


## History

The functions in this package were used in
[Heun et al. [-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109).


## More Information

Find more information, including vignettes and function documentation, at 
<https://MatthewHeun.github.io/matsbyname/>.


## References
