---
title: "Aggregation in `matsbyname`"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    smart: false
vignette: >
  %\VignetteIndexEntry{Aggregation in `matsbyname`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(tidyr)
library(matsbyname)
library(matsindf)
```


## Introduction

The `matsbyname` package provides several
functions to assist renaming and aggregating rows and columns of matrices.
This vignette shows how to use those functions.


## `aggregate_byname()`

By default (`aggregation_map = NULL, margin = c(1,2), pattern_type = "exact"`),
`aggregate_byname()` sums all rows and columns with the same names, 
with the effect that all row and column names are unique.

```{r}
m <- matrix(c(1, 2, 3, 
              4, 5, 6, 
              7, 8, 9), nrow = 3, ncol = 3, byrow = TRUE,
            dimnames = list(c("duck", "duck", "goose"), c("Coal", "Oil", "Solar")))
m
aggregate_byname(m)
```

An `aggregation_map` can be provided,
giving instructions for rows or columns to aggregate and 
the names of the aggregates.
`aggregation_map` should be a list of named strings.
The entries in the list give names of rows and columns to be aggregated.
The names of the list entries provide the names of the resulting aggregates.

```{r}
aggregate_byname(m, aggregation_map = list(birds = c("duck", "goose"), 
                                           `Fossil fuels` = c("Coal", "Oil")))
```

The margin over which the aggregation is to be performed 
is given in the `margin` argument
(1 for rows, 2 for columns). 

```{r}
aggregate_byname(m, aggregation_map = list(`Fossil fuels` = c("Coal", "Oil")), 
                 margin = 2)
```

Note that it is an error to aggregate over a margin and leave 
identically named rows or columns. 
The following function call will fail, 
because it aggregates over both rows and columns,
leaving two "duck" rows.

```{r}
# aggregate_byname(m, aggregation_map = list(`Fossil fuels` = c("Coal", "Oil")))
```



## Pieces

First, understand pieces.


## `rename_to_piece_byname()`







## `aggregate_pieces_byname()`




## Summary

The `matsbyname` package simplifies aggregation of matrix rows and columns
based on row and column names. 
In particular the functions
`aggregate_byname()`, 
`rename_to_piece_byname()`, and 
`aggregate_pieces_byname()`
provide flexibility in how renaming and aggregation can be accomplished.