---
title: "Writing Custom Functions for matsbyname"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    smart: false
vignette: >
  %\VignetteIndexEntry{matsbyname custom functions}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(matsbyname)
```

## Introduction

The `matsbyname` package provides many useful functions for \"by name\" manipulation of
matrices, lists of matrices, and matrices in columns of data frames.
However, the built-in functions may not cover all possible needs.
`matsbyname` provides three functions for these situations:

| Function | Purpose    
|--:|:------------------------------------
| `unaryapply_byname` | apply a unary function to a single matrix, a list of matrices, or a column of matrices in a data frame  
| `elementapply_byname` | apply a unary function to a single element of a matrix or a column of matrices in a data frame
| `binaryapply_byname` | apply a binary function to two matrices or Map a binary function across two lists of matrices or two columns of matrices in a data frame
| `cumapply_byname` | apply a binary function cumulatively to a single list of matrices or a column of matrices in a data frame


## How the `*apply_byname` functions work

The `*apply_byname` functions have several arguments.

| Argument | Description
|--:|:------------------------------
`FUN` | a unary function in the case of `unaryapply_byname`, a binary function (that may also accept a single argument) in the case of `binaryapply_byname`, and a binary function (that must also accept only a single argument) in the case of `cumapply_byname`
`a` | a matrix, a list or matrices, or a column of matrices in a data frame
`b` | a matrix, a list or matrices, or a column of matrices in a data frame
`.FUNdots` | a named list of arguments to be passed to `FUN`
 `rowcoltypes` | tells what to do with row and column types
 `match_type` | tells how row and column types of `a` and `b` arguments must be matched
 `.organize` | tells whether to automatically complete `a` and `b` relative to each other and sort the rows and columns of the completed matrices

`FUN` is mapped as expected over `a` (in the case of `unaryapply_byname`, 
`elementapply_byname`, and `cumapply_byname`) 
or over `a` and `b` (in the case of `binaryapply_byname`).
`FUN` should assume that its `a` and/or `b` arguments are single numbers or matrices; 
`*apply_byname` handles all mapping across lists.
The following sections describe each `*apply_byname` function.


## `unaryapply_byname`

`unaryapply_byname` applies `FUN` to a single matrix, a list of matrices, or 
(if used with `dplyr::mutate()`) a column in a data frame that contains matrices.
The `rowcoltypes` argument must be one of the following:

| `rowcoltypes` value | Behaviour
|--:|:---------------------------------
`"all"` | transfer both row and column types of \code{a} directly to output (the default)
`"transpose"` | rowtype of \code{a} becomes coltype of output; coltype of \code{a} becomes rowtype of output
`"row"` | rowtype of \code{a} becomes both rowtype and coltype of output
`"col"` | coltype of \code{a} becomes both rowtype and coltype of output
`"none"` | rowtype and coltype not set by this function; \code{FUN} will set rowtype and coltype

A simple example follows.

```{r}
U <- matrix(1:4, ncol = 2, dimnames = list(c("p1", "p2"), c("i1", "i2"))) %>%
  setrowtype("Products") %>% setcoltype("Industries")
U
difference_byname(0, U)
unaryapply_byname(`-`, U)
```


## `elementapply_byname`

`elementapply_byname` applies `FUN` to a single matrix, a list of matrices, or 
(if used with `dplyr::mutate()`) a column in a data frame that contains matrices.

A simple example follows.

```{r}
divide <- function(x, divisor){
  x/divisor
}
m <- matrix(c(1:4), nrow = 2, ncol = 2, dimnames = list(c("r1", "r2"), c("c1", "c2"))) %>% 
  setrowtype("row") %>% setcoltype("col")
m
elementapply_byname(divide, a = m, row = 1, col = 1, .FUNdots = list(divisor = 2))
```


## `binaryapply_byname`

`binaryapply_byname` applies `FUN` to a pair of matrices, a pair of lists of matrices, or 
(if used with `dplyr::mutate`) a pair of columns in a data frame that contains matrices.

`match_type` must be one of `"all"`, `"matmult"`, or `"none"`.

| `match_type` value | Behaviour
|--:|:--------------------------------------------
`"all"` | rowtypes of `a` must match rowtypes of `b` and coltypes of `a` must match coltypes of `b` (the default)
`"matmult"` | coltypes of `a` must match rowtypes of `b`
`"none"` | neither coltypes nor rowtypes are checked

The `rowcoltypes` argument (a boolean) 
tells whether to apply row and column types from `a` and `b`
to the output.

The `.organize` argument (a boolean) 
tells whether to automatically complete `a` and `b` relative to each other and
sort the rows and columns of the completed matrices.
Normally, this should be `TRUE` (the default).

A simple example follows.

```{r}
U <- matrix(1:4, ncol = 2, dimnames = list(c("p1", "p2"), c("i1", "i2"))) %>%
  setrowtype("Products") %>% setcoltype("Industries")
U
Y <- matrix(1:4, ncol = 2, dimnames = list(c("p2", "p1"), c("i2", "i1"))) %>%
  setrowtype("Products") %>% setcoltype("Industries")
Y
sum_byname(U, Y)
binaryapply_byname(`+`, U, Y)
```

## `cumapply_byname`

`cumapply_byname` applies `FUN` cumulatively to a list of numbers, 
a list of matrices, or (if used with `dplyr::mutate`) a column of a data frame.
`FUN` must be a binary function that also allows a single argument.
The result is a list with first element `FUN(m[[1]])`.
For `i >= 2`, elements of the resulting list are `FUN(m[[i]], out[[i-1]])`, 
where `out` is the result list.

Simple examples follow.

```{r}
cumapply_byname(sum_byname, list(1, 2, 3, 4))
cumapply_byname(hadamardproduct_byname, list(1, 2, 3, 4))
```


## Summary

The various `*apply_byname` functions allow users to extend the functionality of the `matsbyname` 
package as needed for any problem domain.
The functions are used extensively in `matsbyname` itself.
In fact, all `matsbyname` functions utilize the `*apply_byname` functions,
so they are ready for prime time!