---
title: "Using summarise in matsbyname"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using summarise in matsbyname}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(matsbyname)
library(tibble)
```

## Introduction

`matsbyname` functions in which 
operands are specified in a `...` argument
are ambiguous when applied to a data frame. 
But there is an argument (`.summarise`)
that signals intention, allowing the ambiguous
functions to be used flexibly with data frames.


## "Normal" functions

For normal functions, such as `+` and `mean()`, 
there is no ambiguity about their operation in a data frame.

```{r}
df <- tibble::tribble(~x, ~y, ~z, 
                       1,  2,  3, 
                       4,  5,  6)
# Typically, operations are done across rows.
df %>% 
  dplyr::mutate(
    a = x + y + z,
    b = rowMeans(.)
  )
```

To perform the same operations down columns, 
use `dplyr::summarise()`.

```{r}
df %>% 
  dplyr::summarise(
    x = sum(x), 
    y = sum(y), 
    z = sum(z)
  )
df %>% 
  dplyr::summarise(
    x = mean(x), 
    y = mean(y), 
    z = mean(z)
  )
```


## `matsbyname::sum_byname()`

What does `matsbyname::sum_byname()` mean for a data frame? 
Will it give sums across rows (as `+`), or
will it give sums down columns (as `summarise()`)?
This ambiguity is present for all `*_byname()` functions
in which operands are specified via the `...` argument, 
including `matrixproduct_byname()`, 
`hadamardproduct_byname()`, 
`mean_byname()`, etc.

To resolve the ambiguity, 
use the `.summarise` argument.
The default value of `.summarise` is `FALSE`,
meaning that the functions normally operate across rows.
If you want to perform the action down columns, 
set `.summarise = TRUE`.

```{r}
df %>% 
  dplyr::mutate(
    a = sum_byname(x, y, z), 
    b = mean_byname(x, y, z)
  )
df %>% 
  dplyr::summarise(
    x = sum_byname(x, .summarise = TRUE) %>% unlist(), 
    y = sum_byname(y, .summarise = TRUE) %>% unlist(), 
    z = sum_byname(z, .summarise = TRUE) %>% unlist()
  )
```


## Summary

The `.summarise` argument broadens the range of applicability 
for many `matsbyname` functions, especially when used with data frames.
The default is `.summarise = FALSE`, meaning that operations 
will be performed across columns.
Set `.summarise = TRUE` argument to signal intent
to perform operations down a column.
